Cognate beginnings to bilingual lexical acquisition

### 2023-05-25 manuscript/_appendix/_model-details.qmd
Make site

```{r apafg}
#| label: fig-prior-mean
#| echo: false
#| message: false
#| warning: false
#| fig-width: 7
#| fig-height: 3
#| out-width: 90%
#| fig-cap: "Expected prior-predicted mean."
nd <- expand.grid(n_phon_std = 0,
                  age_std = (10:36-mean(responses$age))/(sd(responses$age)),
                  exposure_std = 0,
                  lv_std = 0)

>>>>

```{r fig-prior-mean}
#| label: fig-prior-mean
#| echo: false
#| message: false
#| warning: false
#| fig-width: 7
#| fig-height: 3
#| out-width: 90%
#| fig-cap: "Expected prior-predicted mean."
nd <- expand.grid(n_phon_std = 0,
                  age_std = (10:36-mean(responses$age))/(sd(responses$age)),
                  exposure_std = 0,
                  lv_std = 0)

### 2023-05-25 manuscript/manuscript.qmd
Make site

format:
  docx:
    reference-doc: "templates/reference.docx"
    knitr:
      opts_chunk:
        dev: "png"
        dev-args:
          - "cairo-png"
        dpi: 300
        out.width: "80%"
  pdf:
    papersize: letter
    fontsize: 12pt
    linestretch: 1.5
    date-format: full
    template-partials:
    - templates/title.tex
    - templates/before-body.tex
    include-in-header:
    - templates/include-in-header.tex
    header-includes: 
    - \usepackage{tipa}
    - \usepackage{nowidow}
    geometry:
    - top=1in
    - bottom=1in
    - left=1in
    - right=1in
    colorlinks: true
    linkcolor: DodgerBlue
    urlcolor: DodgerBlue
    citecolor: DodgerBlue
    link-citations: true
    biblio-config: false
    cite-method: biblatex
    biblatex-chicago: false
    biblio-style: apa
    biblatexoptions: backend=biber
    block-headings: false
    indent: true

@@

Most research on early word acquisition relies extensively on data from monolingual children, and is oblivious to the fact that a substantial proportion of the world population acquires more than one language from early ages [@grosjean2021extent]. Previous research on bilingual vocabulary acquisition pointed to bilingual toddlers knowing, on average, less words in each of their languages than their monolinguals peers, and also to both groups knowing a similar amount of words when the bilinguals' two languages are taken into account. @hoff2012dual found that English-Spanish bilingual toddlers in South Florida (United States) knew less words in English than monolinguals did. Both groups knew a similar total amount of words when both English and Spanish vocabularies were counted together, which substantiates the importance of collecting data on both languages when assessing bilinguals' communicative development. Other studies have provided converging evidence that bilinguals know a similar or even larger number of words than monolinguals when the two languages are aggregated [@oller2002language; @pearson1994patterns; @pearson1993lexical; @patterson2004comparing; @patterson2004bilingual; @smithson2014bilingualism; @petitto2001bilingual; @gonzalez-barrero2020bilinguala]. These studies rely on bilingual samples whose languages are relatively distant in typology: English is a Germanic language, while Spanish is a Romance language, and both languages differ substantially at the grammatical and the phonological level. Little is known about vocabulary acquisition in bilinguals learning a pair of typologically closer languages (e.g., when both are Romance languages).

@@

Not all words occur in the linguistic input with identical frequency, so children accumulate experience with some word-forms at faster rates than with other word-forms. This results in frequent words being acquired earlier in age than less frequent words [e.g., @ambridge2015ubiquity; @goodman2008doesa; @braginsky2019consistency; @swingley2018quantitative; @hansen2017what; @verhagen2022determinants]. For this reason, lexical frequency plays a central role in most models of lexical growth, especially in the case of accumulator models [@hidaka2013computational; @mollica2017how]. Following this account, words with high lexical frequency like 'flower' [`r format(floor(childes[childes$token=="flower",]$freq_million), big.mark = ",")` times per million words in the CHILDES corpus, @macwhinney2000childes] are predicted to be acquired earlier in age than words with low lexical frequency such as 'taxi' (`r format(floor(childes[childes$token=="taxi",]$freq_million), big.mark = ",")` times per million words).

We argue that previous accounts on vocabulary growth based on accumulator models are a convenient framework to study bilingual vocabulary growth. Bilinguals share fundamentally similar learning mechanisms with their monolinguals peers. Developmental divergences between both groups can be understood as the result of the the negotiation between such mechanisms and the particular demands of a dual language input [e.g., @werker2005primir; @curtin2011bilinguala]. One of such divergences is the fact that bilinguals' frequency of exposure to a word-form is mediated by their degree of exposure to their languages. Assuming that bilinguals and monolinguals are exposed to an equivalent amount of linguistic input, a Spanish-English bilingual who is exposed to Spanish 80% of the time and to English 20% of the time would need more time to accumulate the same experience with 'flower' and 'taxi' than an English monolingual (who listens 100% of the time to English) would need.

@@

```{r fig-hypotheses}
#| label: fig-hypotheses
#| fig-cap: "Schematic representation of the hypothesised cognate facilitation effect on word acquisition."
#| echo: FALSE
#| out-width: 100%
#| fig-width: 6
#| fig-height: 5
knitr::include_graphics(normalizePath("assets/img/diagram.png"))
```

@@

To operationalise participants' exposure rate to a word-form, we used a composite measure that captures the lexical frequency of the word-form (an approximation of how often a participant is exposed to it), and child's amount of exposure to the language such word belongs to. This measure was calculated by multiplying the word-form's lexical frequency with participant's degree of exposure to the language the word belongs to. Lexical frequencies were extracted from the CHILDES corpora [@macwhinney2000childes; @sanchez2019childesdb] (see Appendix A for more details). Participants' degree of exposure were reported by caregivers in a language exposure questionnaire they filled before the vocabulary checklist. Caregivers provided a proportion of the time the child was estimated to listen to any language they were spoken to regularly (described in the methods section). The resulting measure is a exposure-weighted lexical frequency measure, which was included in analyses as a proxy of how often bilingual participants were exposed to each word-form.

@@

Finally, we included word length as a nuance predictor in our analyses, motivated by previous studies reporting a strong association between word length and vocabulary size estimates [e.g., @braginsky2019consistency; @jones2019children]. Sorter word-forms like /\textipa{sO\textltilde}/ (Catalan for *sun*) are generally simpler to encode as phonological representations in the lexicon than longer word-forms like /\textipa{ku.ku"DRi\textltilde}/ (Catalan for *crocodile*), and are therefore more likely to be participant a child's vocabulary.

@@

All materials, data, and reproducible code can be found at the OSF ([https://osf.io/hy984/](https://osf.io/hy984/)) and GitHub ([https://github.com/gongcastro/trajectories](https://github.com/gongcastro/trajectories)) repositories. This study was conducted according to guidelines laid down in the Declaration of Helsinki, and was approved by the Drug Research Ethical Committee (CEIm) of the IMIM Parc de Salut Mar, reference 2020/9080/I.

@@

We modelled the probability of participants answering each response category (*No* < *Understands* < *Understands and Says*) using a Bayesian, mixed-effects ordinal regression model. This model allowed us to estimate both item and participant word-acquisition trajectories, while estimating the effect of our variables of interest. We added $\text{Age}$ (number of months elapsed between participants' birth date and questionnaire completion), $\text{Length}$ (number of phonemes in the X-SAMPA phonological transcription of the word-form), $\text{Exposure}$ (composite measure of lexical frequency and language degree of exposure), and $\text{Cognateness}$ (a measure of cognateness, defined as the phonological similarity between translation equivalents) as population-level effects, together with their two- and three-way interactions. Predictors were standardised before entering the model by subtracting the mean of the predictor from each value and dividing the result by the standard deviation of the predictor. Participant-level and item-level random intercepts and slopes were included where appropriate, according to the structure of the data [see @eq-linear, @barr2013random]. We specified a weakly informative prior around the parameters of the model (see Appendix C for details).

@@

It might be argued that our results reflect the fact that cognates are represented in the initial bilingual lexicon as the *same* lexical entry. Because cognates correspond to similar sounding word-forms in equivalent referential contexts (e.g., hearing /\textipa{"gat}/ and /\textipa{"ga.to}/ in the presence of a cat), it is possible that they classify both are as acceptable variations of the same word, therefore treating them as a single lexical item. This would lead to a faster increase in cumulative learning instances, and to earlier ages of acquisition for cognate translation equivalents (for which listening to each word-form contributes to the acquisition of its shared representation), compared to non-cognates (for which listening to each word-form contributes to the acquisition of a separate representation). This mechanism could potentially explain the earlier age-of-acquisition effect of cognates found in the present study, without the need of parallel activation playing any relevant role. Previous work points in a different direction: bilingual toddlers show monolingual-level sensitivity slight phonetic changes in a word-form, according to their performance in word recognition tasks [@swingley200511montholds; @ramon-casas2017minimalpair; @swingley2000spoken; @tamasi2017pupillometry; @mani2011does; @bailey2002phonological; @wewalaarachchi2017vowels; @ramon-casas2010are; @ramon-casas2009vowel], and word learning [@ramon-casas2017minimalpair; @mattock2010first; @havy2016phonetic]. Overall, there is compelling evidence that bilingual toddlers form distinct lexical representations for even near-identical cognates.

>>>>

format:
  pdf:
    papersize: letter
    fontsize: 12pt
    linestretch: 1.5
    date-format: full
    template-partials:
    - templates/title.tex
    - templates/before-body.tex
    include-in-header:
    - templates/include-in-header.tex
    header-includes: 
    - \usepackage{tipa}
    - \usepackage{nowidow}
    geometry:
    - top=1in
    - bottom=1in
    - left=1in
    - right=1in
    colorlinks: true
    linkcolor: DodgerBlue
    urlcolor: DodgerBlue
    citecolor: DodgerBlue
    link-citations: true
    biblio-config: false
    cite-method: biblatex
    biblatex-chicago: false
    biblio-style: apa
    biblatexoptions: backend=biber
    block-headings: false
    indent: true

@@

Most research on early word acquisition relies extensively on data from monolingual children, and is oblivious to the fact that a substantial proportion of the world population acquires more than one language from early ages [@grosjean2021extent]. Previous research on bilingual vocabulary acquisition pointed to bilingual toddlers knowing, on average, less words in each of their languages than their monolinguals peers, and also to both groups knowing a similar amount of words when the bilinguals' two languages are taken into account. @hoff2012dual found that English-Spanish bilingual toddlers in South Florida (United States) knew less words in English than monolinguals did. Both groups knew a similar total amount of words when both English and Spanish vocabularies were counted together, which substantiates the importance of collecting data on both languages when assessing bilinguals' communicative development. Other studies have provided converging evidence that bilinguals know a similar or even larger number of words than monolinguals when the two languages are aggregated [@oller2002language; @pearson1994patterns; @pearson1993lexical; @patterson2004comparing; @patterson2004bilingual; @smithson2014bilingualism; @petitto2001bilingual; @gonzalez-barrero2020bilingual]. These studies rely on bilingual samples whose languages are relatively distant in typology: English is a Germanic language, while Spanish is a Romance language, and both languages differ substantially at the grammatical and the phonological level. Little is known about vocabulary acquisition in bilinguals learning a pair of typologically closer languages (e.g., when both are Romance languages).

@@

Not all words occur in the linguistic input with identical frequency, so children accumulate experience with some word-forms at faster rates than with other word-forms. This results in frequent words being acquired earlier in age than less frequent words [e.g., @ambridge2015ubiquity; @goodman2008does; @braginsky2019consistency; @swingley2018quantitative; @hansen2017what; @verhagen2022determinants]. For this reason, lexical frequency plays a central role in most models of lexical growth, especially in the case of accumulator models [@hidaka2013computational; @mollica2017how]. Following this account, words with high lexical frequency like 'flower' [`r format(floor(childes[childes$token=="flower",]$freq_million), big.mark = ",")` times per million words in the CHILDES corpus, @macwhinney2000childes] are predicted to be acquired earlier in age than words with low lexical frequency such as 'taxi' (`r format(floor(childes[childes$token=="taxi",]$freq_million), big.mark = ",")` times per million words).

We argue that previous accounts on vocabulary growth based on accumulator models are a convenient framework to study bilingual vocabulary growth. Bilinguals share fundamentally similar learning mechanisms with their monolinguals peers. Developmental divergences between both groups can be understood as the result of the the negotiation between such mechanisms and the particular demands of a dual language input [e.g., @werker2005primir; @curtin2011bilingual]. One of such divergences is the fact that bilinguals' frequency of exposure to a word-form is mediated by their degree of exposure to their languages. Assuming that bilinguals and monolinguals are exposed to an equivalent amount of linguistic input, a Spanish-English bilingual who is exposed to Spanish 80% of the time and to English 20% of the time would need more time to accumulate the same experience with 'flower' and 'taxi' than an English monolingual (who listens 100% of the time to English) would need.

@@

```{r fig-hypotheses}
#| label: fig-hypotheses
#| fig-cap: "Schematic representation of the hypothesised cognate facilitation effect on word acquisition."
#| fig-width: 8
#| fig-height: 7
#| out-width: 100%
#| out-height: 100%
#| echo: false
#| fig-align: center
threshold <- 300

# no parallel activation
eli_df <- generate_eli(
    threshold = threshold,
    age = seq(1, 50, length.out = 100),
    freq_month = 50,
    freq_beta = 1,
    l1_doe = 0.60,
    parallel_beta = 0,
    conditions = c(
        glue::glue("**Cognate**: /<i style='color:{clrs[1]}'>ˈgat</i>-<i style='color:{clrs[4]}'>ˈga.to</i>/ [cat]"),
        glue::glue("**Non-cognate**: /<i style='color:{clrs[1]}'>ˈgos</i>-<i style='color:{clrs[4]}'>ˈpe.ro</i>/ [dog]")
    )
) |>
    filter(hypothesis=="H0") |> 
    mutate(language = ifelse(language=="L1", "Catalan form", "Spanish form"))

aoa_df <- generate_aoa(eli_df, threshold = threshold) |>
    mutate(across(aoa, lst(min, max)), .by = c(te, language)) |>
    filter(hypothesis=="H0")
# 
# img <- c(cat = "images/diagram_cat.png",
#          dog = "images/diagram_dog.png") |>
#     map(magick::image_read) |>
#     map(\(x) magick::image_ggplot(x, interpolate = FALSE))

plot_h0 <- eli_df |>
    ggplot(aes(age, eli,
               colour = language)) +
    facet_grid(~ te) +
    geom_segment(data = aoa_df,
                 aes(x = aoa,
                     xend = aoa,
                     y = 0,
                     yend = threshold),
                 linewidth = 3/4) +
    geom_hline(yintercept = aoa_df$threshold,
               linewidth = 3/4,
               colour = "grey80") +
    annotate(geom = "label",
             label = glue::glue("Threshold: {threshold}"),
             label.r = unit(0, "lines"),
             label.size = 0,
             fill = "grey80",
             x = 0,
             y = threshold,
             hjust = 0) +
    geom_line(linewidth = 1) +
    geom_label(data = aoa_df,
               aes(x = aoa,
                   y = threshold,
                   label = round(aoa, 2),
                   fill = language),
               size = 3,
               label.r = unit(0, "lines"),
               label.size = 0,
               colour = "white",
               na.rm = TRUE,
               show.legend = FALSE
    ) +
    labs(x = "Age (months)",
         y = "Cumulative\nlearning instances",
         colour = "",
         shape = "Hypothesis",
         fill = "",
         title = "H0: no parallel activation",
         linetype = "Hypothesis") +
    scale_x_continuous(breaks = seq(min(eli_df$age),
                                    max(eli_df$age), 4)) +
    scale_colour_manual(values = clrs[c(1, 4)]) +
    theme(legend.position = "bottom",
          legend.key.width = unit(1.5, "cm"),
          legend.margin = margin(c(0, 0, 0, 0)),
          legend.justification = c(0, 1),
          legend.title = element_text(size = 9),
          legend.text= element_text(size = 8))

# with parallel activation
eli_df <- generate_eli(
    threshold = threshold,
    age = seq(0, 50, by = 1),
    l1_doe = 0.60,
    freq_month = 50,
    freq_beta = 1,
    conditions = c(
        glue::glue("**Cognate**: /<i style='color:{clrs[1]}'>ˈgat</i>-<i style='color:{clrs[4]}'>ˈga.to</i>/ [cat]"),
        glue::glue("**Non-cognate**: /<i style='color:{clrs[1]}'>ˈgos</i>-<i style='color:{clrs[4]}'>ˈpe.ro</i>/ [dog]")
    )
) |>
    filter(hypothesis=="H1") |> 
    mutate(language = ifelse(language=="L1", "Catalan form", "Spanish form"))

aoa_df <- generate_aoa(eli_df, threshold = 250) |>
    mutate(across(aoa, lst(min, max)), .by = c(te, language)) |>
    filter(hypothesis=="H1")

plot_h1 <- eli_df |>
    ggplot(aes(age, eli,
               colour = language)) +
    facet_grid(~ te) +
    geom_segment(data = aoa_df,
                 aes(x = aoa,
                     xend = aoa,
                     y = 0,
                     yend = threshold),
                 linewidth = 3/4) +
    geom_hline(yintercept = aoa_df$threshold,
               linewidth = 3/4,
               colour = "grey80") +
    annotate(geom = "label",
             label = glue::glue("Threshold: {threshold}"),
             label.r = unit(0, "lines"),
             label.size = 0,
             fill = "grey80",
             x = 0,
             y = threshold,
             hjust = 0) +
    geom_line(linewidth = 1) +
    geom_label(data = aoa_df,
               aes(x = aoa,
                   y = threshold,
                   label = round(aoa, 2),
                   fill = language),
               size = 3,
               label.r = unit(0, "lines"),
               label.size = 0,
               colour = "white",
               na.rm = TRUE,
               show.legend = FALSE
    ) +
    labs(x = "Age (months)",
         y = "Cumulative\nlearning instances",
         colour = "",
         shape = "Hypothesis",
         fill = "",
         title = "H1: parallel activation",
         linetype = "Hypothesis") +
    scale_colour_manual(values = clrs[c(1, 4)]) +
    scale_x_continuous(breaks = seq(min(eli_df$age),
                                    max(eli_df$age), 4)) +
    theme(legend.position = "none")

plot_h0 + plot_h1 +
    plot_layout(ncol = 1, guides = "collect") &
    plot_annotation(tag_levels = "A") +
    scale_colour_manual(values = clrs[c(1, 4)]) &
    scale_fill_manual(values = clrs[c(1, 4)]) &
    scale_y_continuous(labels = function(x) format(x, big.mark = ",")) &
    guides(colour = guide_legend(override.aes = list(linewidth = 2))) &
    theme_ggdist() &
    theme(legend.position = "top", 
          legend.text = element_text(size = 10),
          legend.key.width = unit(2, "cm"),
                    legend.key.height = unit(0.5, "cm"),
          strip.background = element_rect(fill = "grey90", colour = "grey90"),
          strip.text = ggtext::element_markdown(size = 10),
          panel.grid.major.y = element_line(colour = "grey",
                                            linetype = "dotted"),
          plot.background = element_rect(fill = "white",
                                         colour = NA))

@@

To operationalise participants' exposure rate to a word-form, we used a composite measure that captures the lexical frequency of the word-form (an approximation of how often a participant is exposed to it), and child's amount of exposure to the language such word belongs to. This measure was calculated by multiplying the word-form's Zipf-transformed lexical frequency to participant's degree of exposure to the language the word belongs to. Lexical frequencies were extracted from the CHILDES corpora [@macwhinney2000childes; @sanchez2019childesdb] (see Appendix A for more details). Participants' degree of exposure were reported by caregivers in a language exposure questionnaire they filled before the vocabulary checklist. Caregivers provided a proportion of the time the child was estimated to listen to any language they were spoken to regularly (described in the methods section). The resulting measure is a exposure-weighted lexical frequency measure, which was included in analyses as a proxy of how often bilingual participants were exposed to each word-form.

@@

Finally, we included word length (numbwe of phonemes) as a nuance predictor in our analyses, motivated by previous studies reporting a strong association between word length and vocabulary size estimates [e.g., @braginsky2019consistency; @jones2019children]. Shorter word-forms like /\textipa{sO\textltilde}/ (Catalan for *sun*) are generally simpler to encode as phonological representations in the lexicon than longer word-forms like /\textipa{ku.ku"DRi\textltilde}/ (Catalan for *crocodile*), and are therefore more likely to be participant a child's vocabulary.

@@

All materials, data, and reproducible code can be found at the OSF ([https://osf.io/hy984/](https://osf.io/hy984/)) and GitHub ([https://github.com/gongcastro/cognate-beginnings](https://github.com/gongcastro/cognate-beginnings)) repositories. This study was conducted according to guidelines laid down in the Declaration of Helsinki, and was approved by the Drug Research Ethical Committee (CEIm) of the IMIM Parc de Salut Mar, reference 2020/9080/I.

@@

We modelled the probability of participants answering each response category (*No* < *Understands* < *Understands and Says*) using a Bayesian, mixed-effects ordinal regression model. This model allowed us to estimate both item and participant word-acquisition trajectories, while estimating the effect of our variables of interest. We added *Age* (number of months elapsed between participants' birth date and questionnaire completion), *Length* (number of phonemes in the X-SAMPA phonological transcription of the word-form), *Exposure* (composite measure of Zipf-transformed lexical frequency and language degree of exposure), and *Cognateness* (a measure of cognateness, defined as the phonological similarity between translation equivalents) as population-level effects, together with their two- and three-way interactions. Predictors were standardised before entering the model by subtracting the mean of the predictor from each value and dividing the result by the standard deviation of the predictor. Participant-level and item-level random intercepts and slopes were included where appropriate, according to the structure of the data [see @eq-linear, @barr2013random]. We specified a weakly informative prior around the parameters of the model (see Appendix C for details).

@@

It might be argued that our results reflect the fact that cognate translation equivalents are represented in the initial bilingual lexicon as the *same* lexical entry. Because cognates correspond to similar sounding word-forms in equivalent referential contexts (e.g., hearing /\textipa{"gat}/ and /\textipa{"ga.to}/ in the presence of a cat), it is possible that they classify both are as acceptable variations of the same word, therefore treating them as a single lexical item. This would lead to a faster increase in cumulative learning instances, and to earlier ages of acquisition for cognate translation equivalents (for which listening to each word-form contributes to the acquisition of its shared representation), compared to non-cognates (for which listening to each word-form contributes to the acquisition of a separate representation). This mechanism could potentially explain the earlier age-of-acquisition effect of cognates found in the present study, without the need of parallel activation playing any relevant role. @mitchell2022cognates discuss this possibility as a possible explanation of the cognate facilitation effect, in which bilinguals only need to map one word-form to the referent in the case of cognates, while mapping two distinct word-forms in the case of non-cognates. Previous work on mispronunciation perception and minimal pair learning points in a different direction. Bilingual toddlers show monolingual-level sensitivity slight phonetic changes in a word-form, according to their performance in word recognition tasks [@swingley200511montholds; @ramon-casas2017minimalpair; @swingley2000spoken; @tamasi2017pupillometry; @mani2011does; @bailey2002phonological; @wewalaarachchi2017vowels; @ramon-casas2010are; @ramon-casas2009vowel]. These fine abilities to differentiate between similar-sounding word-forms are reflected in word learning, as bilinguals seem to be able to map minimal pairs to distinct referents [@ramon-casas2017minimalpair; @mattock2010first; @havy2016phonetic]. Overall, it seems that bilinguals consider small differences in the phonological forms of words as relevant at the lexical level. We argue that this shows evidence that bilingual toddlers likely form distinct lexical representations for even near-identical cognates.
